---
title: "Project 1 - NFL Summary Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(readr)
library(tidyverse)
```

```{r}
# Read data
nfl <- read_csv("nfl_teams_season_summary.csv")
nfl %>% mutate(division2 = substring(division, 1, 3)) -> nfl
```

Theme: comparison of divisions AFC vs NFC

Research Questions:

1. Are any of the divisions getting stronger/more competitive over time?

Group by division (keeping all 8)
Compare w/ AFC vs NFC (not caring about N/S/E/W)
Is the distribution of wins/losses tight or spread out?
Pull out most competitive teams for each division over the years

```{r}
# Max win/loss ratio from each division
nfl %>%
    mutate(wl_ratio = wins/losses) %>%
    group_by(season, division2) %>%
    slice(which.max(wl_ratio)) %>%
    select(season, team, wl_ratio, division2) %>%
    ggplot(aes(x = season, y = wl_ratio, color = division2)) + 
        geom_line() +
        theme_bw()

library(gridExtra)
# Total wins for each division
4 * sum(nfl$wins)/nrow(nfl)
nfl %>%
    mutate(wl_ratio = wins/losses) %>%
    filter(division2 == "NFC") %>%
    group_by(season, division) %>%
    summarize(total_wins = sum(wins), wl_ratio_avg = mean(wl_ratio)) %>%
    ggplot(aes(x = season, y = total_wins)) +
        geom_hline(yintercept=31) +
        geom_line(color="blue") +
        ylim(21, 42) +
        facet_wrap(~division) +
        theme_bw() -> wins.nfc.plt

nfl %>%
    mutate(wl_ratio = wins/losses) %>%
    filter(division2 == "AFC") %>%
    group_by(season, division) %>%
    summarize(total_wins = sum(wins), wl_ratio_avg = mean(wl_ratio)) %>%
    ggplot(aes(x = season, y = total_wins)) + 
        geom_hline(yintercept=31) +
        geom_line(color="red") +
        ylim(21, 42) +
        facet_wrap(~division) +
        theme_bw() -> wins.afc.plt

library(ggridges)
library(ggthemes)
library(plotly)
nfl %>%
    mutate(wl_ratio = wins/losses) %>%
    ggplot(aes(x=wl_ratio, y=as.factor(season))) + 
        geom_density_ridges() +
        xlim(-1, 5) +
        geom_vline(xintercept=1) +
        theme_pander()

grid.arrange(wins.nfc.plt, wins.afc.plt, nrow = 1)
```


2. Which works better -- passing or running?

Among the winningest teams in each division, how much do they pass vs run? (as a percentage of total yards) 
Wins vs passing yards
Wins vs running yards

```{r}
nfl %>% 
    mutate(total_yds = pass_off_total_yards_gained + run_off_total_yards_gained) %>%
    group_by(season, division2) %>%
    summarise(total_yards = sum(total_yds)) %>%
    ggplot(aes(x = season, y = total_yards, color = division2)) + geom_line() + geom_smooth(method = "lm", se = F)

```


3. Offensive vs defense importance for wins? Which is more impactful?

Look at a sample of the best teams and a sample of the worst teams.

See how many passing/running yards the defense allows.
Look at how many passing/running yards the offense makes.




