---
title: "Project 1 - NFL Summary Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(readr)
library(tidyverse)
```

```{r}
# Read data
nfl <- read_csv("nfl_teams_season_summary.csv")
```

Theme: comparison of divisions AFC vs NFC

Research Questions:

1. Are any of the divisions getting stronger/more competitive over time?

Group by division (keeping all 8)
Compare w/ AFC vs NFC (not caring about N/S/E/W)
Is the distribution of wins/losses tight or spread out?
Pull out most competitive teams for each division over the years


2. Which works better -- passing or running?

Among the winningest teams in each division, how much do they pass vs run? (as a percentage of total yards) 
Wins vs passing yards
Wins vs running yards


3. Offensive vs defense importance for wins? Which is more impactful?

Look at a sample of the best teams and a sample of the worst teams.

See how many passing/running yards the defense allows.
Look at how many passing/running yards the offense makes.




